\chapter*{Acknowledgments}

I first want to acknowledge my mother---without her I wouldn't be here
writing this dissertation.  My father taught me a love of science and
nature, which led to my interest in computers and programming
languages---without him I wouldn't be here, either.  I dedicate my
dissertation to them, with love and admiration.

I also thank my brother Brian, my sister Mary, and their spouses,
Claudia D\'{i}az-Byrd and Donald Stevens.  I can't imagine better
siblings or in-laws.

Renzhong Chen has been part of the Byrd family for twenty years.  I
thank him for his friendship, and for the incredible tour of China that
was a highlight of my time in grad school. I am also grateful for the
hospitality of Renzhong's wife, Lea~Li.

Dan Friedman has been my teacher, mentor, boss, hacking buddy,
coauthor, and friend during my six years at Indiana University.  It
was Dan who introduced me to logic programming, and to his language
that eventually evolved into miniKanren.

I cannot thank Dan without also thanking his wife Mary and the rest of
the Friedmans, who have been a second family to me in Bloomington.  I
am especially grateful to Sa{\tt \{nd|mm\}}i Friedman for hours of
amusement.

My committee members, Dan, Amr, Chris, and Larry, have been
unfailingly cheerful, patient, and supportive.  This is fortunate,
since their intellects might otherwise be intimidating.  I am
especially gratified that all of these scholars are deeply committed
to the art of teaching.  Thank you all!

Oleg Kiselyov taught me an unbelievable amount about logic
programming, especially the benefits of purity, and the dangers of
impure operators like \scheme|conda| and \scheme|condu|.  It was Oleg
who had the incredibly clever idea of implementing a relational
arithmetic system inspired by hardware half-adders and full-adders
(see Chapter~\ref{arithchapter}).  Oleg has also had a tremendous
influence on the development and evolution of miniKanren and its
predecessor, Kanren.

Chung-chieh (Ken) Shan has also greatly influenced the evolution, and
especially the implementation, of miniKanren.  Much of the brevity and
elegance of the core miniKanren implementation in
Chapter~\ref{mkimplchapter} is due to Ken, who designed the critical
\scheme|case-inf| macro.

Some of the ideas, implementation code, and example programs in this
dissertation were first presented, often in a slightly different form,
in {\em The Reasoned Schemer}~\cite{trs}.  Chapter~\ref{arithchapter}
is based on chapters seven and eight of {\em The Reasoned Schemer}
and, to a lesser extent,
\citet{conf/flops/KiselyovBFS08}. Chapters~\ref{akchapter} and
\ref{akimplchapter} are adapted from~\citet{alphamk}.
Chapter~\ref{alphatapchapter} is adapted
from~\citet{DBLP:conf/iclp/NearBF08}.  A paper containing the contents
of Chapter~\ref{fernschapter} and the nestable engines code in
Appendix~\ref{nestable-engines} will be presented at Mitch Wand's
Festschrift.  Many thanks to all of my coauthors.  Please see the
acknowledgments sections of these papers for additional credits.

The first tabling implementation for miniKanren was designed by the
author and Ramana Kumar, and inspired by the Dynamic Reordering of
Alternatives (DRA) approach to
tabling~\cite{dra09,simpleimplementingtabling}.
Ramana implemented the design, with debugging assistance from the
author.  The tabling implementation presented in Chapter~\ref{tablingchapter}
is a slightly modified version of Ramana's second, and improved,
tabling implementation.  Most importantly, the new implementation is
based on streams rather than success and failure continuations, which
means answers are produced in the same order as in the core miniKanren
implementation of Chapter~\ref{mkimplchapter}.

The nominal unifier using triangular substitutions in
section~\ref{triangularsection} is due to Joe Near.  Ramana Kumar
has implemented a faster but very different triangular unifier.

As described in section~\ref{walkexpensive}, Abdulaziz Ghuloum, David
Bender, and Lindsey Kuper have explored which purely functional data
structures are best for representing triangular substitutions.

The \scheme|pmatch| pattern-matching macro in Appendix~\ref{pmatch}
was written by Oleg Kiselyov.  The \scheme|matche| and
\scheme|lambdae| pattern-matching macros in Appendix~\ref{matche} were
originally designed by the author and implemented by Ramana Kumar with
the help of Dan Friedman.  Andy Keep, Michael Adams, and Lindsey Kuper
worked with us to implement an optimized version of \scheme|matche|
and \scheme|lambdae|, which will be presented at the 2009 Scheme
Workshop.

Visits to Bloomington from Christian Urban, Matt Lakin, and Gopal
Gupta greatly aided my research.  I also benefited from the 3rd
International Compulog/ALP Summer School on Logic Programming and
Computational Logic at New Mexico State University, organized by
Enrico Pontelli, Inna Pivkina, and Son Cao Tran.  I enjoyed many
stimulating conversations with visiting scholars Juliana Vizzotto,
Katja Grace, Dave Herman, and Sourav Mukherjee.

Indiana University's PL Wonks lecture series, organized by Roshan
James, has been most stimulating.  I thank Roshan, Michael Adams, Andy
Keep, Jeremiah Willcock, Ron Garcia, Jeremy Siek, Steve Ganz, Larisse
Voufo, and all the other Wonks for many interesting conversations
about programming languages.  The PL Wonks also benefited from special
visits by Jeffrey Siskind and Robby Findler.  I look forward to a
relaxing conversation with Olivier Danvy that does not require either
of us to use our first-responder skills.

For the past eleven semesters I have had the great pleasure of being
the associate instructor for Indiana University's undergraduate (C311)
and graduate (B521) introductory programming languages courses, under
the enthusiastic leadership of Dan Friedman.  The material in this
dissertation was greatly improved by the comments and corrections of
our students.  I am grateful to you all, especially those former
students who have conducted summer research with us: Dave Bender,
Jordan Brown, Adam Foltzer, Adam Hinz, Andy Keep, Jiho Kim, Ramana
Kumar, Lindsey Kuper, Micah Linnemeier, and Joe Near.

Special thanks goes to former C311 student Jeremiah Penery, who
discovered and corrected a subtle error in the definition of
\scheme|logo| from the first printing of {\em The Reasoned Schemer}.

Dan and I have had the great fortune to work with exceptional graduate
and undergraduate associate instructors: David Mack, Alex Platte, Kyle
Blocher, Joe Near, Ramana Kumar, and Lindsey Kuper.  Their hard
work has made C311 and B521 such a success.  Joe and Ramana also read
the final draft of this dissertation, and provided many insightful
comments and corrections---thank you!

I learned a great deal from teaching the honors section of IU's
introductory programming course (C211) under supreme Scheme hacker
Kent Dybvig.  Several years later I had the great pleasure of teaching
C211 with another Scheme master, Aziz Ghuloum, as my associate
instructor.  Every teacher should be so lucky.

I thank Olin Shivers for writing an exemplary dissertation, the
structure and organization of which I shamelessly ripped off.  Once
again I relied on Dorai Sitaram's Scheme typesetting program, S\LaTeX.

Whenever I was floundering in grad school Mitch Wand seemed to
magically appear, pulling me aside to see how I was doing, and
offering much appreciated advice and encouragement.  Although I did
not always follow his advice (to my detriment, I am sure), I will
always be grateful for his support.

In addition to sharing his teaching and programming expertise, Kent
Dybvig also offered invaluable advice about navigating the pitfalls of
grad school.  I didn't heed all of Kent's advice, though I learned to
pay special attention to anything following the catchphrase, ``You'll
be committing academic suicide.''

Lucy Battersby, Rob Henderson, and the rest of the Indiana University
Computer Science Department staff provided expert help and an
outstanding work environment.

The friendly staff of the east-side Bloomington Quiznos, Sunny Bal,
Kae Lunde, Caitlyn Muncy, Alisha Findley, Tylla Carlisle, and Meagan
Perry, kept me rolling in veggie subs and ``om-nom-nom-nom''-worthy
cookies.

Caitlin Coar, Cortney Packett, and Alisha Stout, formerly of Cold
Stone Creamery, supplied me with delicious and nourishing PB\&C
milkshakes.

I wrote most of this dissertation at the east-side Starbucks in
Bloomington, where the generous, hilarious, and slightly-unhinged
baristas provided a tasty setting for extended writing sessions.  Many
thanks to:

\begin{itemize}
\item Amanda Buck for all the fish stories;
\item Andrea Jerabek for always giving me a hard time;
\item Ben Canary for inventing the infamous and delectable ``Christmas Cookie'';
\item Brian Ibison for adoring Olivia Munn;
\item Christina Liwski for never giving me a hard time;
\item Ciera Brannon for not believing in cells (\emph{No cells, no mercy!});
\item Eric ``Big Eric'' Martin for pointing out that UFO over Starbucks;
\item Erin Dobias for being a smart chick;
\item Gabby Baehl for living up to her first name;
\item Megan Traxinger-James for putting up with Eric and Brian;
\item and Phil Wood for his Twitter-powered news reports.
\end{itemize}

\noindent I also thank former baristas Jessica Fugate and Shannon
Pilrose for hanging out with me when I should have been writing.

% Chili's ???  Marnie, Todd, Sara,

I've been fortunate to have made many close friends in Bloomington.
In particular, Aziz Ghuloum, Larisse Voufo, Ron Garcia, Suzanna Crage,
Andy Keep, Lindsey Kuper, Lindsay and Ahmed Hamed, and Anne and Mike
Faber helped keep me sane while I wrote this dissertation.

Marc Muher visited Bloomington for Dig Dug \emph{battle royale}.
Leslie Cuevas helped keep Jennifer Fitzgerald in line, while Ada
Brunstein offered end game encouragement.  As always, my childhood
friends Mike, Daryl, and Bobby helped relieve the tension with the
occasional game.

It has been my honor to know the Miller family for many years, and am
grateful for their friendship.

Cisco Nochera, former director of Camp Greentop, has been my friend
and mentor for almost two decades.

During the spring of 2006 I visited Chile for a week, along with my
brother Brian and my sister-in-law Claudia.  Claudia's parents, Hugo
D\'{i}az and Nancy Zu\~{n}iga de D\'{i}az, warmly welcomed us to their
home in the beautiful countryside on the outskirts of Casablanca,
Chile.  I thank them for their never-ending hospitality and goodwill.

I have had many incredible teachers in my life, but a few stand out.
My absolutely amazing 11$^{th}$ grade Spanish teacher, Tom Rahauser,
taught me to never fear Se\~{n}or Subjunctivo.  Richard Saenz's class
on special relatively completely blew my mind.  Tom Anastasio taught
me LISP, cleverly disguised as C.  Alan Sherman prepared me for
graduate school.  Dan Friedman taught me that you don't understand
your code until it fits on a 3x5 card.
