\begin{tabular}{l l}
\hspace{-0.92cm}
\begin{minipage}{2.5in}
\thispagestyle{empty}
William E. Byrd

Dept. of Computer Science

Lindley Hall 215

Indiana University

Bloomington, IN 47405

{\tt webyrd@cs.indiana.edu}

(812) 855-4885

\end{minipage}
\hspace{2cm}
\begin{minipage}{2.5in}
\vspace{-1.35cm}
home:

3488 E. Covenanter Dr.

Bloomington, IN 47401

(812) 320-8505
\end{minipage}
\end{tabular}

\bigskip

\noindent{\bf Degrees}

\medskip

\noindent \ \ B.S. in Computer Science, 1999, University of Maryland Baltimore County,  

\noindent \hspace{1cm} Baltimore, Maryland, {\it cum laude}

\medskip

\noindent \ \ B.S. in Special Education, 1994, College of Charleston, Charleston, 

\noindent \hspace{1cm} South Carolina, {\it magna cum laude}

\bigskip

\noindent{\bf Current Position}

\medskip

\noindent \ \ Assistant Instructor under the direction of Dan Friedman.

\bigskip

\noindent{\bf Honors}

\medskip

\noindent \ \ Benefitfocus.com ``Medal of Honor'', Benefitfocus.com, 2001.

\medskip

\noindent \ \ AppNet Excellence Award, AppNet, 2000.

\medskip

\noindent \ \ Outstanding Senior in Computer Science and Electrical Engineering, 

\noindent \hspace{1cm} University of Maryland, Baltimore County, 1999.

\bigskip

\noindent{\bf Books}

\medskip

\noindent\ \ Friedman, D. P., Byrd, W. E., and Kiselyov, O. {\it The Reasoned Schemer},

\noindent \hspace{1cm} The MIT Press, 2005.

% \bigskip

% \noindent{\bf Publications}

\bigskip

\noindent{\bf Conferences}

\medskip

\noindent\ \ Near, J., Byrd, W. E., and Friedman, D. P. ``\alphatap: A Declarative 

\noindent \hspace{1cm} Theorem Prover for First-Order Classical Logic'', In Proceedings of the

\noindent \hspace{1cm}  24th International Conference on Logic Programming, volume 5366 of

\noindent \hspace{1cm}  Lecture Notes in Computer Science, pp. 238--252, 2008.

\medskip

\noindent\ \ Kiselyov, O., Byrd, W. E., Friedman, D. P., and Shan, C. ``Pure, Declarative, 

\noindent \hspace{1cm} and Constructive Arithmetic Relations (Declarative Pearl)'', In Proceedings 

\noindent \hspace{1cm}  of the 9th International Symposium on Functional and Logic Programming, 

\noindent \hspace{1cm}  volume 4989 of Lecture Notes in Computer Science, pp. 64--80, 2008.

\bigskip

\noindent{\bf Workshops}

\medskip

\noindent\ \ Byrd, W. E. and Friedman, D. P. ``\alphakanren: A Fresh Name in Nominal

\noindent \hspace{1cm} Logic Programming'', In Proceedings of the 2007 Workshop on Scheme 

\noindent \hspace{1cm} and Functional Programming, Universit\'{e} Laval Technical Report 

\noindent \hspace{1cm} DIUL-RT-0701, pp. 79--90, 2007.

\medskip

\noindent\ \ Byrd, W. E. and Friedman, D. P. ``From Variadic Functions to Variadic

\noindent \hspace{1cm} Relations'', In Proceedings of the 2006 Scheme and Functional

\noindent \hspace{1cm} Programming Workshop, University of Chicago Technical Report 

\noindent \hspace{1cm} TR-2006-06, pp. 105--117, 2006.


%\bigskip

%\bigskip

%*** home info should be right flush, at same height as work info ***

%*** Are right-hand margins correct?  They don't look wide enough ***

%*** Remember to fill out publications section if ICLP or POPL papers are accepted ***
